\documentclass{amsart}
\usepackage{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{xifthen}
\usepackage{tikz-cd}

% Theorems
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{remark}[thm]{Remark}
\newtheorem{defn}[thm]{Definition}

% Math stuffs
\newcommand{\pow}[1]{\ensuremath{\mathcal{P}(#1)}}
\newcommand{\card}[1]{\ensuremath{\left\vert #1 \right\vert}}
\newcommand{\real}{\ensuremath{\mathbb{R}}}
\newcommand{\nat}{\ensuremath{\mathbb{N}}}
\newcommand{\cbult}{\ensuremath{\mathrm{CBUlt}}}
\newcommand{\cle}{\ensuremath{\lesssim}}
\newcommand{\ceq}{\ensuremath{\cong}}
\newcommand{\relR}{\ensuremath{\mathrel{R}}}
\newcommand{\relS}{\ensuremath{\mathrel{S}}}
\newcommand{\AND}{\ensuremath{\mathrel{\wedge}}}
\newcommand{\OR}{\ensuremath{\mathrel{\vee}}}
\newcommand{\op}{\ensuremath{\mathsf{op}}}
\newcommand{\den}[1]{\ensuremath{\llbracket #1 \rrbracket}}
\newcommand{\defs}{\ensuremath{\mathrel{\triangleq}}}
\newcommand{\bifix}{\ensuremath{\mathsf{bifix}}}

% Sets
\newcommand{\term}{\ensuremath{\mathit{Term}}}
\newcommand{\urel}{\ensuremath{\mathit{URel}}}

% judgments
\newcommand{\guardJ}[2]{\ensuremath{\text{$#1.\,#2$\ \textsf{guarded}}}}
\newcommand{\hasE}[2]{\ensuremath{#1 \mathrel{:} #2}}
\newcommand{\hasKJ}[2]{\ensuremath{#1 \vdash \hasE{#2}{\mathrm{\kind}}}}
\newcommand{\hasTJ}[5]{\ifthenelse{\isempty{#1}}%
  {\ensuremath{#3 \vdash \hasE{#4}{#5}}}%
  {\ensuremath{#1;#3 \vdash \hasE{#4}{#5}}}}
\newcommand{\hasEJ}[6]{\ifthenelse{\isempty{#1}}%
  {\ensuremath{#3; #4 \vdash \hasE{#5}{#6}}}%
  {\ensuremath{#1; #3; #4 \vdash \hasE{#5}{#6}}}}
\newcommand{\hasCEJ}[9]{\ensuremath{#6; #7; #8 \vdash #1 : #2; #3; #4%
    \vdash #5 \rightsquigarrow #9}}
\newcommand{\hasCJ}[7]{\ensuremath{\vdash #1 : \cdot; #2; #3 \vdash #4%
    \rightsquigarrow \cdot; #5; #6 \vdash #7}}
\newcommand{\equivJ}[6]{\ensuremath{#2; #3 \vdash #4 \sim #5 : #6}}
\newcommand{\equivSubst}[4]{\ensuremath{#4 \vdash #1 \sim #2 : #3}}
\newcommand{\step}[2]{\ensuremath{#1 \mapsto #2}}
\newcommand{\steps}[2]{\ensuremath{#1 \mapsto^* #2}}
\newcommand{\dec}[4][i]{\ensuremath{#2 \rhd^{#1} #3 : #4}}
\newcommand{\valueJ}[1]{\ensuremath{#1\ \mathsf{value}}}
\newcommand{\subKJ}[3]{\ifthenelse{\isempty{#1}}%
{\ensuremath{{#2} \leq {#3}}}%
{\ensuremath{{#1} \vdash {#2} \leq {#3}}}}

% language
\newcommand{\kind}{\ensuremath{\mathsf{kind}}}
\newcommand{\later}{\ensuremath{{\blacktriangleright}}}
\newcommand{\rec}[2]{\ensuremath{\mu #1.\, #2}}
\newcommand{\fn}[2]{\ensuremath{#1 \to #2}}
\newcommand{\tp}{\ensuremath{\mathsf{T}}}
\newcommand{\unit}{\ensuremath{\mathsf{unit}}}

\newcommand{\ap}[2]{\ensuremath{#1\ #2}}
\newcommand{\lam}[3]{\ensuremath{\lambda #1 {:} #2.\, #3}}
\newcommand{\into}[1]{\ifthenelse{\isempty{#1}}%
  {\ensuremath{\mathsf{in}}}%
  {\ensuremath{\mathsf{in}\,#1}}}
\newcommand{\out}[1]{\ifthenelse{\isempty{#1}}%
  {\ensuremath{\mathsf{out}}}%
  {\ensuremath{\mathsf{out}\,#1}}}
\newcommand{\delay}[1]{\ifthenelse{\isempty{#1}}%
  {\ensuremath{\mathsf{next}}}%
  {\ensuremath{\mathsf{next}\,#1}}}
\newcommand{\letdelay}[3]{\ensuremath{\mathsf{let\ next\ } #1 = #2 \mathsf{\ in\ } #3}}
\newcommand{\all}[3]{\ensuremath{\forall #1 {:} #2.\, #3}}
\newcommand{\allNoKind}[2]{\ensuremath{\forall #1.\, #2}}

\newcommand{\Ap}[2]{\ensuremath{#1[#2]}}
\newcommand{\Lam}[3]{\ensuremath{\Lambda #1 {:} #2.\, #3}}

\newcommand{\zap}{\ensuremath{\circledast}}


\title{The Next 700 Failed Step-Index-Free Logical Relations}
\author{Daniel Gratzer}
\date{\today}

\begin{document}
\begin{abstract}
  An important question in programming languages is the study of
  program equivalence. This is done typically with the construction of
  a relational denotational model or a syntactic analogue, called a
  logical relation. Logical relations have proven to be an effective
  tool for analyzing programs and lending formal weight to ideas like
  data abstraction and information hiding. A central difficulty with
  logical relations is their fragility; it has proven to be a
  challenge to scale logical relations to more realistic languages. In
  this thesis, we discuss a number of methods for the construction of
  a logical relation for a language with general references. None of
  these methods are sufficient to define a logical relation without
  resorting to step-indexing, demonstrating the difficulty of
  expressing the recursive structure of higher-order references.
\end{abstract}
\maketitle

In the study of programming languages a great number of important
questions can be reduced to questions about the equality of
programs. The verification of a compiler pass is nothing but a
question of equality of the naive program and its optimized version,
an optimized data structure may be shown to be correct relative to a
much simpler reference solution, a type theory hinges on the
construction of definitional equality and so on. Given the role that
equality plays in programming languages, a great variety of
mathematical tools have been developed to analyze the various notions
of equality possible in a programming language.

When in this thesis equality is discussed, it is generally meant to be
\emph{contextual equivalence}. Contextual equivalence is formally
defined later (see
Definition~\ref{defn:language:contextualEquivalence}) but informally
contextual equivalence of two programs, $e_1 \cong e_2$, expresses
that $e_1$ and $e_2$ are internally indistinguishable. This means that
there is no program containing $e_1$ so that if $e_1$ is replaced with
$e_2$ then the program returns a different result. This notion of
equivalence is appealing because there is no aspect of a program we
care about other than how it computes. Contextual equivalence allow to
ignore unimportant differences in the precise way that the answer of a
program is computed and focus instead on the answer itself. This means
that two well-behaved implementations of a data structure, for
instance, are contextually equivalent even if one is far more complex
and efficient than the other.

For all the appeal of contextual equivalence, it is very difficult to
establish that two programs are contextually equivalent. In order to
establish that $e_1 \cong e_2$ it is necessary to quantify over all possible
programs using $e_1$ and $e_2$ and reason about their behavior. This
includes programs which do not really make use of $e_1$ or $e_2$ such
as $\ap{(\lam{x}{\fn{\unit}{\tau}}{1})}{\lam{x}{\unit}{e_1}}$. This
lack of constraints on how $e_1$ or $e_2$ is used is precisely what
makes contextual equivalence so useful but it makes even the most
basic proofs involved affairs. In order to compensate for this, a
variety of tools have been developed to simply the process of
establishing when $e_1 \cong e_2$.

Broadly speaking, there are two classes of these tools. One may
consider denotational approaches, where in the equality of programs is
reduced to the question of the equality of normal mathematical
objects. This line of study begins with Dana Scott's investigations of
the lambda calculus~\citep{TODO-CITE-SCOTT}. These tools have been
immensely effective when they may be applied but they are difficult to
use. Complex programming languages often make use of extremely
sophisticated mathematical objects and using the model requires
understanding them. This has meant that denotational semantics is
traditionally out of reach for the verification of programs by an
average programmer or anyone besides a domain expert.

On the other hand, there are syntactic approaches to equality. Some of
these date back to the original study of the lambda calculus and its
reduction properties. Syntactic tools tend to be simpler to use,
relying only one elementary mathematics and an understanding of the
syntax itself. For equality, the tool of choice when working
syntactically is a logical relation~\citep{TODO-TAIT-AMAL}. Dating
back to~\citet{TODO-TAIT} logical relations crucially define a
type-indexed notion of equality. Logical relations have proven
important for validating certain obviously desirable properties such
as function extensionality. They have also given rise to a powerful
theory of data abstraction called
parametricity~\citep{TODO-REYNOLDS}. Parametric reasoning as inspired
by logical relations has been so important that there have even been
attempts to reconstruct it in a purely abstract denotational
sense~\citep{TODO-FIBRATIONAL}.

To be precise, a logical relation in for us is a family of sets
indexed by the types of the language, $(R_\tau)_\tau$. It is
constructed by induction over the types indexing it so that
$R_{\fn{\tau_1}{\tau_2}}$ is defined in terms of $R_{\tau_1}$ and
$R_{\tau_2}$. For each $\tau$, the following property is expected to
hold:
\[
  (e_1, e_2) \in R_\tau \implies
  \hasE{\cdot}{e_1, e_2}{\tau} \land e_1 \cong e_2
\]
This property expresses the soundness of the logical relation with
regards to contextual equivalence: logical equivalence implies
contextual equivalence. The reverse property, completeness, is
desirable but often unachievable\footnote{Instead a variety of
  techniques for \emph{forcing} completeness to hold are
  employed. These amount to adding all the missing identifications to
  the logical relation. While theoretically desirable, this is useless
  for the problem of actually establishing an
  equivalence}. Completeness is usually unnecessary for establishing
certain concrete equivalences so in this context it is less
important. Finally, an important property of logical relations which
is difficult to capture formally is that it is much easier to prove
that $(e_1, e_2) \in R_\tau$ than to directly show that
$e_1 \cong e_2$. Typically, $R_\tau$ is meant to capture the logical
action of $\tau$; it expresses the precise set of observations
possible to make of expressions of type $\tau$ without all the
duplication of contextual equivalence. For instance, in order to show
that $(e_1, e_2) \in R_{\fn{\tau_1}{\tau_2}}$ it suffices to show the
following.
\[
  \forall (a_1, a_2) \in R_{\tau_1}.
  \ (\ap{e_1}{a_1}, \ap{e_2}{a_2}) \in R_{\tau_2}
\]
Given the obvious appeal of having a logical relation defined for a
language, it would seem that the construction of a logical relation
characterizing equality is a natural first step in the study of a new
programming language. The central stumbling block to this goal is that
logical relations are difficult to extend to new programming languages
and especially to new types and computational effects.

For an instance of where trouble might arise, consider a language with
polymorphic types: $\allNoKind{\alpha}{\tau}$. What should the
definition of $R_{\allNoKind{\alpha}{\tau}}$

\input{secs/introduction}
\input{secs/language}
\input{secs/step-indexing}
\input{secs/domains}
\input{secs/handedness}
\input{secs/guarded-recursion}
\input{secs/conclusion}

\bibliographystyle{plainnat}
\bibliography{citations}
\end{document}
